---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, echo=FALSE}
library(ggplot2)
library(data.table)
```

## Download the CSV file from online URL

```{r  echo=FALSE}
file <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(file, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'), method = "curl")
unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")

```

## Now read the downloaded file in to a DataTable variable

```{r echo=FALSE}
activityDT <- data.table::fread(input = "data/activity.csv")
head(activityDT)
```

## Calculate total no of steps taken each day. This is calculated by taking standard deviation

```{r echo=FALSE}
total_steps_pd <- activityDT[, c(lapply(.SD, sum, na.rm = FALSE)),  by = .(date), .SDcols = c("steps")] 
```

## Now plot the toal_steps_pd using ggplot- histogram
```{r echo=FALSE}
ggplot(total_steps_pd, aes(x = steps)) +
    geom_histogram(fill = "blue", binwidth = 200) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

## Calculate mean and median of total no of steps taken each day
```{r echo=FALSE}
total_steps_pd[, .(mean_steps = mean(steps, na.rm = TRUE), median_steps = median(steps, na.rm = TRUE))]
```

## 5-minute interval that, on average, contains the maximum number of steps
```{r echo=FALSE}
intervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval)] 
```

## Plot time series plot for average number of steps taken
```{r echo=FALSE}
ggplot(intervalDT, aes(x = interval , y = steps)) + geom_line(color="blue", size=1) + labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")
```


## Imputing NA values from the dataTable

```{r echo=FALSE}
# Devise a method to fill missing data
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
```


```{r echo=FALSE}
# Now create a new data table with all the missing values filled in
data.table::fwrite(x = activityDT, file = "data/tidyData.csv", quote = FALSE)

```
### Total steps taken per day
```{r echo=FALSE}
Total_Steps <- activityDT[, c(lapply(.SD, sum)), .SDcols = c("steps"), by = .(date)] 
```

### mean and median total number of steps taken per day
```{r echo=FALSE}
Total_Steps[, .(Mean_Steps = mean(steps), Median_Steps = median(steps))]
```

```{}

Important: Here you can see that mean and median has changed with and without NA values in the data set
```

## Difference between activities between weekday and weekends

```{r echo=FALSE}
activityDT <- data.table::fread(input = "data/activity.csv")
activityDT[, date := as.POSIXct(date, format = "%Y-%m-%d")]
activityDT[, `Day of Week`:= weekdays(x = date)]
activityDT[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
activityDT[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
activityDT[, `weekday or weekend` := as.factor(`weekday or weekend`)]
head(activityDT, 10)
```

## Plotting for Weekend and weekdays activity patterns

```{r echo=FALSE}
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)] 

ggplot(IntervalDT , aes(x = interval , y = steps, color=`weekday or weekend`)) + geom_line() + labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") + facet_wrap(~`weekday or weekend` , ncol = 1, nrow=2)
```